SOM result tables

### 2025-03-01 analysis/Supplementary/supplementary_tables.qmd
Macbook daily work session. Results in main document corrected and rewritten for emphasis on main patterns and comparison between groups. Supplementary document revised and corrected.

solutrean_dataset <- full_join(solutrean_tech, shelter_RM_WIP, by = "ID") # The complete dataset has 2818 (44 points) entries
solutrean_dataset <- solutrean_dataset |>
  drop_na(RMTYPE) |>   # Dropping NAs on the RM database drops dataset to 1514
  drop_na(CLASS) |>  # Dropping NAs on the technological class drops dataset to 914
  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
    RETOUCHEDPIECETYPOLOGY == "Bifacial fragment" ~ "Bifacial blank",
    RETOUCHEDPIECETYPOLOGY == "Shouldered point fragment" ~ "Shouldered point",
    TRUE ~ RETOUCHEDPIECETYPOLOGY))

@@

@tbl-general-gravettian shows the number and percentages of technological classes per chert type. Local cherts show similar patterns between them, with T2 and T5 being the types with higher numbers of artefacts (n=110 and n=72, respectively). For all local cherts, blanks (including blank fragments) represent more than 50% of the artefacts. Cores and core maintenance products are only present in T2 and T5, although in small frequencies: for T2, cores represent 13.6% (n=15 including core fragments) of the group and core preparation and maintenance products represent 1.8% (n=2). For T5 only 4 core fragments and 1 core preparation and maintenance product are present, all frequencies between \~5-1%. Retouched tools seem to be especially relevant in T2 (n= 20, \~18% including retouched piece fragments), making them the most frequent class after blanks. T5 shows similar percentages of retouched tools and shatter (\~12% and \~14%, respectively). T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).

@@

As seen in table @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type. From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group. From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%). Regarding typologies, 15 types of retouched tools were identified (excluding retouched tool fragments). All these typologies were found in the non-local chert group, while only 7 retouched tool typologies were identified in local cherts/chalcedony. The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26). From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7. Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7). The presence of burin spalls in local chert, especially T2 (n=9) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).

@@

The final lithic sample from the Proto-Solutrean assemblage, consisting of the levels 4E and top of level 5, consists of 635 artifacts. Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type. Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert (n=158 or `r round(158*100/(32+62+32+158+51), 2)`% of the local types). Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).

When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%). Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~9-12% of the classes in each local type, with the exception of T1 (chalcedony), with no cores. Core preparation and maintenance products are absent in all local cherts. For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively). Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%). For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site; all other non-local types show at least 1 core and frequently the presence of shatter. As expected, the TL artifacts include a blank and a retouched. However, a shatter was also identified.

@@

In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto). A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts. As seen in @tbl-retouch-proto, 14 different types of retouched tools were identified. Endscrappers (26.2%), splintered pieces (17%), burins (12.3%) and retouched flakes (7.7%) were the most prevalent. Local raw materials show 9 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers. Non-local cherts show 10 different typologies of retouched tools. Perforator-endscrapper, truncation, backed bladelet and retouched blades typologies can only be found in non-local varieties, albeit also in small numbers.

@@

The Solutrean assemblages show similar patterns to those observed in previous assemblages. The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts. Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group. Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 (n=506). Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).

When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, especially T2, the majority of the debitage is composed of flakes and elongated blanks (83.9%). Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%). In comparison to debitage products, retouched tools are scarce in the sample (\~3%). Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%). Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%). However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).

@@

# Blank-to-core ratios

```{r}
#| label: ratio-blankcore

# Solutrean ratio
solutrean_ratiobc <- solutrean_dataset %>% 
  mutate(CLASS = case_when(
    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
    TRUE ~ CLASS)) %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    TRUE ~ CLASS)) %>% 
  filter(RMTYPE != "INDET") %>% 
  filter(CLASS %in% c("Blank", "Core")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Blank/Core" = Blank / Core)

# Proto-Solutrean ratio
proto_ratiobc <- phase_dataset %>% 
  filter(Level %in% c("4E", "4E/5")) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "Core")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>%
  mutate("Blank/Core" = Blank / Core)

# Gravettian ratio
gravettian_ratiobc <- phase_dataset %>% 
  filter(Level %in% c(6, 7)) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "Core")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Blank/Core" = Blank / Core)

@@

# Tool-to-debitage ratios

```{r}
#| label: ratios-retouchdebitage

# Solutrean
solutrean_ratiotd <- solutrean_dataset %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
    CLASS == "Retouched piece" ~ "RetouchedPiece",
    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
    TRUE ~ CLASS)) %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Tool/Blank" = RetouchedPiece / Blank)

# Proto-Solutrean
proto_ratiotd <- phase_dataset %>% 
  filter(Level %in% c("4E", "4E/5")) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Tool/Blank" = RetouchedPiece / Blank)

# Gravettian
gravettian_ratiotd <- phase_dataset %>% 
  filter(Level %in% c(6, 7)) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Tool/Blank" = RetouchedPiece / Blank)

@@

# Measurements

```{r}
#| label: stats-setup

@@

# Tables with individual class. For supplementary information if necessary
# For flakes (complete)
solutrean_stats_blank <- solutrean_stats |>  
  filter(CLASS == "Blank") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  drop_na() |> 
  group_by(RMTYPE, CLASS) |> 
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
# For cores
solutrean_stats_cores <- solutrean_stats |> 
  filter(CLASS == "Core") |>
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  drop_na() |> 
  group_by(RMTYPE, CLASS) |> 
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))

@@

# Tables with individual class. For supplementary information if necessary
# For flakes (complete)
proto_stats_blank <- proto_stats |> 
  filter(CLASS == "Blank") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
# For cores
proto_stats_cores <- proto_stats |> 
  filter(CLASS == "Core") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))

@@

# Tables with individual class. For supplementary information if necessary
# For flakes (complete)
gravettian_stats_blank <- gravettian_stats |> 
  filter(CLASS == "Blank") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
# For cores
gravettian_stats_cores <- gravettian_stats |> 
  filter(CLASS == "Core") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))

@@

```{r}
#| label: tbl-solutrean
#| tbl-cap: "Stats for Solutrean occupation."
#| tbl-subcap: 
#|  - "Stats for blanks"
#|  - "Stats for cores"
#| layout-nrow: 4

>>>>

solutrean_dataset <- full_join(solutrean_tech, shelter_RM_WIP, by = "ID") # The complete dataset has 2818 (44 points) entries
solutrean_dataset <- solutrean_dataset |>
  drop_na(RMTYPE) |>   # Dropping NAs on the RM database drops dataset to 1514
  drop_na(CLASS) |>  # Dropping NAs on the technological class drops dataset to 914
  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
    RETOUCHEDPIECETYPOLOGY == "Bifacial fragment" ~ "Bifacial blank",
    RETOUCHEDPIECETYPOLOGY == "Shouldered point fragment" ~ "Shouldered point",
    RETOUCHEDPIECETYPOLOGY == "Retouched blade" ~ "Retouched blank",
    TRUE ~ RETOUCHEDPIECETYPOLOGY))

@@

@tbl-general-gravettian shows the number and percentages of technological classes per chert type. For local cherts, blanks (including blank fragments) represent more than 50% of the artefacts. Cores and core maintenance products are only present in small frequencies: <10% and <2%, respectively. Retouched tools seem to be especially relevant in local T2-5 cherts (~16%), making them the most frequent class after blanks and blank fragments, followed by shatter (~10%). T1 (chalcedony) shows similar patterns to other local cherts, although with higher percentages of shatter (34.3%).

@@

As seen in table @tbl-retouch-gravettian, 170 retouched tools were identified, with 39 (`r round(39*100/170, 2)`%) having an unidentified chert type. From the identified chert types, `r 3+20+4+2+9` (`r round((3+20+4+2+9)*100/170, 2)`%) of retouched tools are from local types, `r 58+24+3+1+3` (`r round((58+24+3+1+3)*100/170, 2)`%) from non-local cherts and 4 are from the TL group. From all chert types, T6 shows the highest numbers of retouched pieces (n=58, `r round(58*100/170, 2)`%). Regarding typologies, 9 types of retouched tools were identified (excluding retouched tool fragments). All these typologies were found in the non-local chert group, while only 5 retouched tool typologies were identified in local cherts/chalcedony. The most common typologies identified in the assemblage (and for all chert types) were burins (n=52), endscrappers (n=17), retouched flakes (n=30) and splintered pieces (n=26). From all these, the majority were produced in non-local chert, mainly T6 and in lesser numbers T7. Although in lesser numbers as well, backed bladelets were also identified (n=6), but only present in non-local chert types (T6 and T7). The presence of burin spalls in local chert, especially T2 (n=9) and non-local cherts (T6 with n=12 and T7 with n=4) is expected due to the prevalence of burins (@tbl-general-gravettian).

@@

The final lithic sample from the Proto-Solutrean assemblage (levels 4E and top of level 5), consists of 635 artifacts. Of these, 167 (`r round(167*100/635, 2)`%) have an indeterminate chert type. Local cherts correspond to `r 32+62+32+158+51` artefacts (`r round((32+62+32+158+51)*100/635, 2)`%), with T4 being the most present local chert as seen in previous works (Belmiro et al., 2025). Non-local types are present in smaller numbers (n=`r 60+47+13+9+1` or `r round((60+47+13+9+1)*100/635, 2)`%), with T6 and T7 being the most relevant (n=60 and n=47, respectively), and the TL group includes 3 artefacts, corresponding to 2 types of TL cherts (Table s++ in SOM ++).

When crossing the lithic class with chert type (@tbl-general-proto), we see that the local types show high frequency of blanks (\>50%, including blank fragments), followed by shatter (\~11-27%). Cores are present in small numbers (n=`r 4+5+13+6`), corresponding to ~10% of the classes in local types, with the exception of T1 (chalcedony), with no cores. Core preparation and maintenance products are absent in all local cherts. For non-local types, especially those with higher number of artefacts (T6-T8), blanks and blank fragments are the most frequent class (48.3%, 53.1% and 69.2%, respectively). Retouched tools are present in smaller frequencies for the previous mentioned chert types (\~17-23%). For T6 the absence of cores but the presence of shatter and a single core preparation and maintenance product suggests knapping on site; all other non-local types show at least 1 core and frequently the presence of shatter. As expected, the TL artifacts include a blank and a retouched. However, a shatter was also identified.

@@

In total, 65 chert retouched tools were identified, with 23.1% (n=15) having an unidentified chert type (@tbl-retouch-proto). A total of `r 1+10+2+8+5` (`r round((1+10+2+8+5)*100/65, 2)`%) retouched tools correspond to local cherts, while `r 10+9+3+1+1` (`r round((10+9+3+1+1)*100/65, 2)`%) correspond to non-local and TL cherts. As seen in @tbl-retouch-proto, 11 different types of retouched tools were identified. Endscrappers (n=18), retouched blanks (n=14), splintered pieces (n=11) and burins (n=8) were the most prevalent. Local raw materials show 8 different typologies, with Vale Comprido points, Solutrean retouch and notches being present only in the local types, even if in small numbers. Non-local cherts show 8 different typologies of retouched tools. Perforator-endscrapper, truncation and backed bladelet typologies can only be found in non-local varieties, albeit also in small numbers.

@@

The Solutrean assemblages show similar patterns to those observed in previous assemblages. The final sample from levels A, B and C from the Solutrean occupations of the Shelter area consists of 914 artifacts. Of these, 201 (`r round(201*100/914, 2)`%) correspond to the indeterminate chert group. Local cherts correspond to `r 16+506+4+10+30` (`r round((16+506+4+10+30)*100/914, 2)`%) artefacts, mostly from T2 as seen in previous works (Belmiro et al., 2025). Non-local cherts (including the TL group) correspond to `r 86+9+30+3+1+17+1` (`r round((86+9+30+3+1+17+1)*100/914, 2)`%) artefacts, mostly from T6 (n=86).

When crossing the class data with type of raw material (@tbl-general-solutrean), we see that for local chert, the majority of the debitage is composed of blanks (83.9%). Cores are also present (10.5%), alongside a small percentage of core preparation and maintenance products (1.3%). In comparison to debitage products, retouched tools are scarce in the sample (\~3%). Non-local cherts, especially T6 and T8, show similar patterns to the local cherts, with high percentages of blanks (\>50%). Cores and core preparation and maintenance are also present in small percentages (\~3-6% and 3.5% only for T6, respectively), as well as retouched tools (\<10%). However, these specific results are truncated by the lack of shatter, which were not present in the sample used for the study (see Materials and Methods section).

@@

# Blank-to-core ratios
```{r}
#| label: ratio-blankcore

# Solutrean ratio
solutrean_ratiobc <- solutrean_dataset %>% 
  mutate(CLASS = case_when(
    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
    TRUE ~ CLASS)) %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    TRUE ~ CLASS)) %>% 
  filter(RMTYPE != "INDET") %>% 
  filter(CLASS %in% c("Blank", "Core")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "T11", "TL"))) |>
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Blank/Core" = Blank / Core)

# Proto-Solutrean ratio
proto_ratiobc <- phase_dataset %>% 
  filter(Level %in% c("4E", "4E/5")) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "Core")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>%
  mutate("Blank/Core" = Blank / Core)

# Gravettian ratio
gravettian_ratiobc <- phase_dataset %>% 
  filter(Level %in% c(6, 7)) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "Core")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Blank/Core" = Blank / Core)

@@

# Tool-to-debitage ratios
```{r}
#| label: ratios-retouchdebitage

# Solutrean
solutrean_ratiotd <- solutrean_dataset %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "Flake" | CLASS == "ElongBlank" | CLASS == "Thinning flake" ~ "Blank",
    CLASS == "FlakeFrag" | CLASS == "ElongBlankFrag" ~ "BlankFrag",
    CLASS == "Retouched piece" ~ "RetouchedPiece",
    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
    TRUE ~ CLASS)) %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "T11", "TL"))) |>
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Tool/Blank" = RetouchedPiece / Blank)

# Proto-Solutrean
proto_ratiotd <- phase_dataset %>% 
  filter(Level %in% c("4E", "4E/5")) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
   mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Tool/Blank" = RetouchedPiece / Blank)

# Gravettian
gravettian_ratiotd <- phase_dataset %>% 
  filter(Level %in% c(6, 7)) %>% 
  filter(RMTYPE != "INDET") %>% 
  mutate(CLASS = case_when(
    CLASS == "BlankFrag" & PIECECOMPLETENESS == "Proximal" ~ "Blank", # To add proximal fragments to ratio
    CLASS == "RetouchedPieceFrag" & PIECECOMPLETENESS == "Proximal" ~ "RetouchedPiece",
    TRUE ~ CLASS)) %>% 
  filter(CLASS %in% c("Blank", "RetouchedPiece")) %>% 
  count(RMTYPE, CLASS, sort = TRUE) %>% 
  group_by(RMTYPE, CLASS) %>%
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2", "T3", "T4", "T5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(n=sum(n)) %>% 
  spread(CLASS, n, fill=0) %>% 
  mutate(sum = rowSums(across(where(is.numeric)))) %>% 
  filter(!sum <= 5) %>% 
  mutate("Tool/Blank" = RetouchedPiece / Blank)

@@

# Measurements
```{r}
#| label: stats-setup

@@

# Tables with individual class. For supplementary information if necessary
# For flakes (complete)
solutrean_stats_blank <- solutrean_stats |>  
  filter(CLASS == "Blank") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  drop_na() |> 
  group_by(RMTYPE, CLASS) |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
                                        "T8", "T9", "T10", "T11", "TL"))) |>
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
# For cores
solutrean_stats_cores <- solutrean_stats |> 
  filter(CLASS == "Core") |>
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  drop_na() |> 
  group_by(RMTYPE, CLASS) |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
                                        "T8", "T9", "T10", "T11", "TL"))) |>
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))

@@

# Tables with individual class. For supplementary information if necessary
# For flakes (complete)
proto_stats_blank <- proto_stats |> 
  filter(CLASS == "Blank") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
# For cores
proto_stats_cores <- proto_stats |> 
  filter(CLASS == "Core") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))

@@

# Tables with individual class. For supplementary information if necessary
# For flakes (complete)
gravettian_stats_blank <- gravettian_stats |> 
  filter(CLASS == "Blank") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))
# For cores
gravettian_stats_cores <- gravettian_stats |> 
  filter(CLASS == "Core") |> 
  select(RMTYPE, CLASS, MAXWIDTH, LENGTH, WEIGHT) |> 
  group_by(RMTYPE, CLASS) |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("T1", "T2-5", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  summarise(across(c(MAXWIDTH, LENGTH, WEIGHT), list(mean = mean, sd = sd)))

@@

```{r}
#| label: tbl-solutrean
#| tbl-cap: "Stats for Solutrean occupation."
#| tbl-subcap: 
#|  - "Stats for blanks"
#|  - "Stats for cores"
#| layout-nrow: 2
